Scaling up a GF grammar for free parsing

### 2011-12-14 doc/report/report.tex
report: structuring grammar part

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
\usepackage{tipa}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\subsection*{Post-nominal articles} 
Unlike most of the world's languages Swedish express not only the number but also
the definitness of a noun by suffixes. The endings are decided by the noun's gender,
%which may be 
\textit{neuter} or \textit{common}.\\
%(or \textit{utrum}) 
\begin{tabular}{l|lll}
&\textbf{Definite}& \textbf{Indefinite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt} & {katten} & \textit{common}\\
& \emph{a cat} & \emph{the cat} &\\
\textbf{Plural} & katter & katterna &\\
 &\emph{cats} & \emph{the cats} &\\
 \hline
\textbf{Singular} & {ett hus} & {huset} & \textit{neuter}\\
 & \emph{a house} & \emph{the house} & \\
\textbf{Plural} & hus & husen & \\
& \emph{houses} & \emph{the houses} & \\
\end{tabular}\\

@@

\chapter{Importing Saldo}
\label{sec:prog.saldo}
Saldo is compatible with GF, which makes it easy to transform the morphological
lexicon into the GF format. The lexicon has been imported earlier (Angelov,2009)
but the code needed updates to not miss many important words.
Experiments have been made to use a interactive tool for lexical acquisition,
but the dictionary was still much to small. The acquisition tool may still
be used for complementing the lexicon.

\section{Implementation}
%>> syntactic information too. Mention that GF does not have this, only lexical or
%>> semantic. Not 1-1.
The basic algorithm for importing Saldo was implemented by Angelov (2009?)
and  produces code for a GF lexcion.
For each word in Saldo, it decides which forms should be used as input
to the the GF smart paradigms. For a verb, this will mean giving
the present form. \\

\begin{figure}[h]
\verb-mkV "knyter" ;-
\caption{First output for the irregular? verb `knyta'}
\label{fig:saldoknyt}
\end{figure}

@@

Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (eg. the 
compounding form ``väx-")
while the one generated in GF contains some that are not in Saldo (eg. ``vuxits").
Saldo may also contain variats: ``växt" and ``vuxit" are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generates one variant whenever alternatives are given. 
The program continues to generate new lexicons as long as there are more
paradigms to choose from and compare the results after each run.

The program has been made more robust than the previous version, it prints
logs and finally lists of the words that were not imported. The logs provides
information about which paradigms that were used for each word and the result
from comparing the inflection tables.
During the porting of Saldo, some renaming of the lemmas was needed.
Special characters like \emph{å,ä,ö} should not occur in GF function names.
Those are translated into \emph{aa,ae,oe} respectively. 
The Saldo identifier \verb|äta..vb| is thus turned into the GF identifier \verb|aeta_V|.
Since there might be two lemmas that share the same
identifier after the translation, a number is added to the name whenever a translation
have been done: 
\begin{tabular}{lll}
kältisk & $\rightarrow$ & kaeltisk\_1 \\
kaeltisk & $\rightarrow$ & kaeltisk \\
entrecôte &$\rightarrow$ & entrecoote\_1 \\ 
\end{tabular}

@@

%see notes to add more about this, pronouns and the importing itself
%choices for adjectives vs. verbs for particips
The importing can be redone anytime.

@@

\chapter{The grammar}
\label{sec:prog.grammar}
We can't expect to get a grammar giving full coverage, instead we aim for a
grammar that gives
a deep analysis while having a satisfactory coverage.
We have been developing a grammar fragment which have been adapted to cover
constructions used in Talbanken. The starting point has been the Swedish
resource grammar and the new grammar should still be compatible with this.
%A description of the Swedish implementation of the resource grammar is given
%in the next section, while
Before describing what has been added, will give a short introductions to the
the cats of resource grammar and the sw implementation in particular.
This is done in section \ref{sec:swegf} while section \ref{sec:Added} shows and
explains the development during this project.
%Explain what's in the resources, constructions present in most languages. 
%Been working on a grammar fragment based on Talbanken,
%like cool people like Montague. May also be interesting for other purposes.
%Swedish shares with Nor and Dan. Extra module, covering things like sw is a
%v2-lang, focusing parts of sentences.

\section{The Swedish resource grammar}
>> start by easy examples, jag ser inte dig, du ser inte mig osv. 
>> plocka från svenska delen 1 2 4 
>> show simple trees
>> explain all in Sw. section.
>> show abstract and concrete tree, compare. (or in Gf sect)
>> no subjects or object, analyse is more from the lexicon,
>> compare to others, ask elisabet.
\label{sec:swegf}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other Scandinavian languages.
Apart form describing morphology, the resource grammar deals with
word order, agreement, tense, basic conjunction etc.

In figure \ref{fig:gftree1} we saw how the sentence ``Jag ser dig" is parsed.
Discuss questions, how are word order handled here? Maybe negation and anterior.
Let's look on how post-determination (sec ref) is implemented, we know that the
smart paradigms can handle the noun declination, now show 'katten' vs. 'den trötta katten'.
A N carries the inflection information, turned in to CN by setting number. Example, type
of N and CN. A CN may then be modified by APs, set field for this. When turned into NP, 
definitness is set, depends on the isMod -field. NP has case as a parameter.
\verb|DetNP : Det -> CN -> NP ; |
To turn in into a NP need to fix determination, by Det. Dets have .. fixed,
but allow variation in.. 
To form \emph{`fågeln'}, the determiner (DetQuant DefArt NumSg) is used.
Is a quantifier, hence does not require number, is given later.
%What is the difference?
In gf, only personal pronouns are considered to belong to PN. The rests (example from
saldo, which is pronoun in SAG as well) are either Quantifiers or Dets.
Difference, how the determinern can set the definiteness of the noun and
the adjective.
Once we have a NP, can use Predets. What they are.\\

@@

Since two out of the three Scandinavian languages share this tempus, it was 
added to the Scandiniavin extra module.
"jeg kommer til å savne deg"

A table of result can be overview in figure \ref{fig:kommeratt}
\begin{figure}[h]
\begin{verbatim}
s SFutKommer Simul Pos Main : jag kommer att se henne
s SFutKommer Simul Pos Inv  : kommer jag att se henne
s SFutKommer Simul Pos Sub  : jag kommer att se henne
s SFutKommer Simul Neg Main : jag kommer inte att se henne
s SFutKommer Simul Neg Inv  : kommer jag inte att se henne
s SFutKommer Simul Neg Sub  : jag inte kommer att se henne
s SFutKommer Anter Pos Main : jag kommer att ha sett henne
s SFutKommer Anter Pos Inv  : kommer jag att ha sett henne
s SFutKommer Anter Pos Sub  : jag kommer att ha sett henne
s SFutKommer Anter Neg Main : jag kommer inte att ha sett henne
s SFutKommer Anter Neg Inv  : kommer jag inte att ha sett henne
s SFutKommer Anter Neg Sub  : jag inte kommer att ha sett henne
\end{verbatim}
\caption{Parse tree for ``Har han inte ätit de gula äpplena idag".}
\label{gfSwe:parsetree}
\end{figure}

kommer att - for scand, extra tense, example
example table

@@

%% Clauses
\subsubsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite[19]{SAG} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.
Transitive verbs may not be used like this
>> elisabet disagrees: use det äter en fågel ost på taket
\enumsentence{\shortexnt{6}
{*Det & gillar & en & fågel & på & taket}
{It & likes & a & bird & on & the roof"}}

unless their in passive form
\enumsentence{\shortex{5}
{Det&dricks&mycket & öl& nuförtiden}
{It&drink+s&much & beer& nowadays}
{A lot of beer is being drunk these days}}
A very common example of this is sentences with the verbs \emph{finnas},
\emph{saknas} and \emph{fattas}. > engelska.
\enumsentence{\shortex{3}
{Det & finns & kaffe}
{It & exist & coffee}
{''There is coffee"}}
\enumsentence{\shortex{3}
{Det & saknas&  kaffe}
{It & misses & coffee}
{''There is no coffee"}}
\enumsentence{\shortex{3}
{Det & fattas & kaffe}
{It & misses & coffee}
{''There is no coffee"}}

A special GF category, \verb|SimpleVP|, was needed for verb phrases of the type
described above.
There are also restrictions on the subject.
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågeln på taket. & *Det sitter fågeln på taket.\\
``There are that bird sitting on the roof'' & ``There are the bird sitting on the roof'' \\
\end{tabular}}
>> depends on the determiner, must require a DIndef? noun! (Cooper,1986)
The noun must be in indefinite form, even though some determiners are allowed.
\enumsentence{Det sitter några fåglar på taket\\
              ``There are some birds on the roof''}
\enumsentence{*Det sitter fåglarna på taket\\
``There are the birds on the roof''}
%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
The GF grammar does not make any distinction between indefinite and definite noun
phrases. Therefore this is handled by looking at the determiner. \\
\verb|FormalSub : SimpleVP -> Det -> CN -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determined form, the clause is put to NONEXIST.
However, this is not an ideal solution. There are other examples of when we
need to know the definiteness of the noun phrase, for example when using
the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
or when using some predeterminers (see section `Quantifiers'). Not true, not the same :(

@@

\textbf{Genitive} -problem 1 cannot be alone (fixed) 2 some words have id. (no) gen form (fixed)
           fixing this include restructuring all???
\textbf{Det-utr}. not effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection{Reflexive pronouns}
%varandra and sin/sitt/sig.
An important piece in a Swedish grammar is the reflexive pronouns and
%both \emph{sin} and , eg. \emph{sin}.
the possessive pronouns as described in section \ref{swe:refl}.
Those make up noun phrases which are not allowed in subjects, 
and we want to be able to parse to following sentences:

>> remove stupid examples, add 'hon och sin vän åkte tåg'
>> use self instead of his.
\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw his children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
His gloves, he had forgotten on the train.}
\enumsentence{Han ger sina pengar till sina syskon.\\
He gives his money to his children.}
\enumsentence{Han är längre än sin kompis.\\
He is taller than his friend.} \label{ex:ReflAP}
\vspace{-2mm}
\enumsentence{Han är här oftare än sin kompis.\\
He is here more often than his friend.} \label{ex:ReflAdv}
\vspace{-2mm}
\enumsentence{Hon tyckte om skolan och alla sina elever.\\
She liked the school and all her students.}\label{ex:ReflConj}
\enumsentence{Han såg sina få böcker och sin penna.\\
He saw his few books and his pencil.}\label{ex:ReflConj2}
\vspace{-2mm}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all her friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave.}
but not
\enumsentence{*Sina vantar var kvar på tåget.\\
Her gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give his money to his siblings. }
\enumsentence{*Han vill att jag rakar sig.\\
He wants me to shave himself.}

@@

>> allowed by the presence of a restrictive relative 
>> only definite article (determinative pronoun), not determiner
The grammar has been extended to accept
\enumsentence{\shortex{7}
{\textbf{De uppfattningar}& som& förs& fram ...} 
{The opinions& that& put & forward ...}
{``The opinions presented ..."}}
where the definite artcile \emph{``de"} is used together with a indefinite
noun. This is only correct for the definite articles and not determiners
and requires that the noun phrase is followud by a restrictive
relative clauses or together with the verb \emph{``finnas"}.
In other cases definitev form must be used, cf. sentence \ref{sent:debocker}.
\enumsentence{\textbf{De uppfattningarna} förs fram.\\
``The opinions are presented."}\label{sent:debocker}
% also ok in finnas!!

@@

@@

\section{Future Work}
\label{sec:future}
>> divide into consequences and future plans.
What is still left to do? Description and direction of future work:
Handling unknown words and , handling unknown grammar, 
how to make the parser robust, ellipses and long sentences, names and numbers.
What is still needed in the grammar, how important that is. 
\subsection{Grammar}
köpa bil 
object shift
\subsection{Parallel parsing/Chunk parsing}
\subsection{Lexicon}
compounds
%Speed up!
%The list of not-imported can be used for adding other kinds of lexicons - idioms, have pos-tags.
%Also to analyse what's wrong, or use the words with a word guesser. The code
%has been adapted for this.
%For exapmle pronouns - dets, have combined information from Talbanken to add them in the correct
%category. How this works, look at tags. Small amount, can be done manually.

>>>>

\documentclass{report} 
\usepackage[utf8]{inputenc}
\usepackage{ amssymb }
\usepackage{graphicx}
\usepackage[margin=1.5in]{geometry}
\usepackage{lingmacros}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{wrapfig}
\usepackage{tipa}
% exempel: 2) Inte vet jag
%              not know I
%             I don't know

@@

\subsection*{Post-nominal articles} 
\label{sec:swedishnoun}
Unlike most of the world's languages Swedish express not only the number but also
the definitness of a noun by suffixes. The endings are decided by the noun's gender,
%which may be 
\textit{neuter} or \textit{common}.\\
%(or \textit{utrum}) 
\begin{tabular}{l|lll}
&\textbf{Definite}& \textbf{Indefinite}& \textbf{Gender} \\
 \hline
\textbf{Singular} & {en katt} & {katten} & \textit{common}\\
& \emph{a cat} & \emph{the cat} &\\
\textbf{Plural} & katter & katterna &\\
 &\emph{cats} & \emph{the cats} &\\
 \hline
\textbf{Singular} & {ett hus} & {huset} & \textit{neuter}\\
 & \emph{a house} & \emph{the house} & \\
\textbf{Plural} & hus & husen & \\
& \emph{houses} & \emph{the houses} & \\
\end{tabular}\\

@@

\chapter{Importing Saldo}
\label{sec:prog.saldo}
The lexicon provided with the GF resources is far too small for open-domain parsing.
Experiments have been made to use an interactive tool for lexical acquisition,
%but the dictionary was still much to small. The acquisition tool may still
but this should be used for complementing rather than creating the lexicon.
This section describes the process of importing Saldo, which is 
compatible with GF, and easily translated to GF format.
As Saldo is continously updated, the importing process has been designed to be fast
and stable enough to be redone any time.

\section{Implementation}
%>> syntactic information too. Mention that GF does not have this, only lexical or
%>> semantic. Not 1-1.
The basic algorithm for importing Saldo was implemented by Angelov (2009?)
and  produces code for a GF lexcion.
For each word in Saldo, it decides which forms should be used as input
to the the GF smart paradigms. For a verb, this will in most cases mean giving
the present form. \\

\begin{figure}[h]
\verb-mkV "knyter" ;-
\caption{First code produced for the verb `knyta'}
\label{fig:saldoknyt}
\end{figure}

@@

Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (eg. the 
compounding form ``väx-")
while the one generated in GF contains some that are not in Saldo (eg. ``vuxits").
Saldo may also contain variats: ``växt" and ``vuxit" are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generates one variant whenever alternatives are given. \\
%The program continues to generate new lexicons as long as there are more
%paradigms to choose from and compare the results after each run.

The program has been made more robust than the previous version. It also prints
log files providing
information about the process of importing each word: which paradigms that have been
tried, the results from the comparisons of the inflection tables and finally listings
of the words that could not be imported.

Each entry in Saldo has an identifier, eg. \verb-äta..vb-, which are used as 
constant names in the GF lexicon. The identifiers given i Saldo may contain
special characters that should be avoided in constants in GF. The importation therefore
needed some renaming. The Swedish letters \emph{å,ä,ö} are translated into
\emph{aa,ae,oe} respectively. 
\begin{verbatim}äta..vb  -> aeta_V \end{verbatim}
The translation may cause two or more lemmas to share the same
GF identifier, and to avoid name collition a number is added to the name
whenever a translation have been done: 
\begin{tabular}{lll}
kältisk & $\rightarrow$ & kaeltisk\_1 \\
kaeltisk & $\rightarrow$ & kaeltisk \\
entrecôte &$\rightarrow$ & entrecoote\_1 \\ 
\end{tabular}\\

@@

%see notes to add more about this, pronouns and the importing itself
%choices for adjectives vs. verbs for particips

@@

\chapter{The grammar}
\label{sec:prog.grammar}
An important part of this project has been to developed a grammar and
adapting it to cover
constructions used in Talbanken. A grammar can never be expected to give
full coverage of a language, but we aim for a grammar fragment which gives a deep
analysis of the most important Swedish constructions.
The starting point has been the GF resource grammar and our new grammar is
still compatible with this.
Before describing the work progress we will give an introduction to the
resource grammars in general and to the Swedish implementation in particular.
This is found in section \ref{sec:swegf} while section \ref{sec:Added} shows and
explains the development during this project.

%We can't expect to get a grammar giving full coverage, instead we aim for a
%grammar that gives
%a deep analysis while having a satisfactory coverage.
%The starting point has been the Swedish
%resource grammar and the new grammar should still be compatible with this.
%%A description of the Swedish implementation of the resource grammar is given
%%in the next section, while
%Before describing what has been added, will give a short introductions to the
%the cats of resource grammar and the sw implementation in particular.
%This is done in section%Explain what's in the resources, constructions present in most languages. 
%Been working on a grammar fragment based on Talbanken,
%like cool people like Montague. May also be interesting for other purposes.
%Swedish shares with Nor and Dan. Extra module, covering things like sw is a
%v2-lang, focusing parts of sentences.

\section{The Swedish resource grammar}
>> start by easy examples, jag ser inte dig, du ser inte mig osv. 
>> plocka från svenska delen 1 2 4 
>> show simple trees
>> explain all in Sw. section.
>> show abstract and concrete tree, compare. (or in Gf sect)
>> no subjects or object, analyse is more from the lexicon,
>> compare to others, ask elisabet.\\
\label{sec:swegf}
The GF resource grammars contains a fundamental description of Swedish,
covering the morphology and many phrase and clause constructions. Most of the implementation
is shared with the other Scandinavian languages.
Apart form describing morphology, the resource grammar deals with
word order, agreement, tense, basic conjunction etc.

Section \ref{gf:swe} contained an easy example with nouns. We used the categories \verb-N-
and \verb-N'- to separated between nouns with repsectively without a set number.
The full analysis includes a separation between the categories \verb-N- -- simple nouns --
and 
\verb-CN-, common noun phrases which may be modified by adjectival phrases or conjucted.
A \verb-CN- may be turned into a \verb-NP- by the rule \verb-DetNP- which sets
the number and definitness. Since
all morphological information is generated by the smart paradigms and contained in the \verb-N-
the construction of a \verb-NP- more or less consists of putting the pieces together.
The two basic articles are \verb-IndefArt- and \verb-DefArt- which
uses the post-modified form of the noun: katten.
However, we need to consider the rules described in section \ref{sec:swedishnoun}. If we have
a common noun phrase consisting of the parts \emph{`liten'} (\emph{`small'}) and \emph{`katt'}
(\emph{`cat'}), there are three ways they can be combined:
\begin{verbatim}
DetNP : Det -> CN -> NP ;
        en    liten katt     -- DIndef
        den   lilla katten   -- DDef Def
        min   lilla katt     -- DDef Indef
\end{verbatim}        
Which combination that shoud be used is decided by the determiner. Hence all determiners
in our grammar must keep information of which definitness they allow. They record field
has a field storing this and the parameters are called \verb-DIndef-, 
\verb-DDef Indef- and \verb-DDef Def-. The determiners have arguments telling
which number the noun phrase should be in.
We also need to make sure that we do not allow constructions as given in
\ref{sent:trottkatt}a.
The definite article  \emph{``en"}is considered to be a quantifier, in plural it
is the either \emph{``de"} or nothing:
\begin{verbatim}
DetNP : Det              -> CN        -> NP ;
        definite, plural +  katt      =  katter
        definite, plural +  stor katt =  de stora katterna
\end{verbatim}        
This difference depends on the common noun phrase, which must carry information
about whether it has been modified or not.
The resource grammar have differenciate between quantifiers -- \verb-Quant-,
determiners \verb-Det- and predeterminers. Predeterminers modify \verb-NP-s
while the other to modifies \verb-CN-s. The differences are shown in table
\ref{tab:detquant}
\begin{figure}[h]
\begin{tabular}{l|lll}
                        &\textbf{Has number}& \textbf{Has definitness}& Example  \\ 
\hline
\textbf{Predeterminers} & --  & --  & alla katter, all maten \\ 
\textbf{Quantifiers}&-- &\checkmark  & min katt, mina katter, *min katten \\ 
\textbf{Determiners}   &\checkmark &\checkmark & varje katt, *varje katten  \\ 
\end{tabular}
\caption{}\label{tab:detquant}
\end{figure}

In gf, only personal pronouns are considered to belong to PN. The rests (example from
saldo, which is pronoun in SAG as well) are either Quantifiers or Dets.

%In figure \ref{fig:gftree1} we saw the GF representation of the sentence ``Jag
%ser dig".
%Discuss questions, how are word order handled here? Maybe negation and anterior.
%Let's look on how post-determination (sec ref) is implemented, we know that the
%smart paradigms can handle the noun declination, now show 'katten' vs. 'den trötta katten'.
%A N carries the inflection information, turned in to CN by setting number. Example, type
%of N and CN. A CN may then be modified by APs, set field for this. When turned into NP, 
%definitness is set, depends on the isMod -field. NP has case as a parameter.
%\verb|DetNP : Det -> CN -> NP ; |
%To turn in into a NP need to fix determination, by Det. Dets have .. fixed,
%but allow variation in.. 
%To form \emph{`fågeln'}, the determiner (DetQuant DefArt NumSg) is used.
%Is a quantifier, hence does not require number, is given later.
%What is the difference?
%Difference, how the determinern can set the definiteness of the noun and
%the adjective. Ref (Cooper,1986)
%Once we have a NP, can use Predets. What they are.\\

@@

Since two out of the three Scandinavian languages share this tense, it was 
added to the Scandiniavin extra module.
"jeg kommer til å savne deg"

A table of result can be overview in figure \ref{fig:kommeratt}
\begin{figure}[h]
\begin{verbatim}
s SFutKommer Simul Pos Main : jag kommer att se henne
s SFutKommer Simul Pos Inv  : kommer jag att se henne
s SFutKommer Simul Pos Sub  : jag kommer att se henne
s SFutKommer Simul Neg Main : jag kommer inte att se henne
s SFutKommer Simul Neg Inv  : kommer jag inte att se henne
s SFutKommer Simul Neg Sub  : jag inte kommer att se henne
s SFutKommer Anter Pos Main : jag kommer att ha sett henne
s SFutKommer Anter Pos Inv  : kommer jag att ha sett henne
s SFutKommer Anter Pos Sub  : jag kommer att ha sett henne
s SFutKommer Anter Neg Main : jag kommer inte att ha sett henne
s SFutKommer Anter Neg Inv  : kommer jag inte att ha sett henne
s SFutKommer Anter Neg Sub  : jag inte kommer att ha sett henne
\end{verbatim}
\caption{The covered and accepted usage of future modal verb 'komma'}
\label{gfSwe:parsetree}
\end{figure}

@@

%% Clauses
\subsubsection{Impersonal constructions}%Presentering, , Topicalisering
\label{sec:Formal}
Formal subjects \cite[19]{SAG} is often used in Swedish.
\enumsentence{\shortex{6}
{Det & sitter & en & fågel & på & taket}
{It & sits & a & bird & on & the roof}
{``There is a bird sitting on the roof''}}
\emph{`Det'} has the position of the subject, and the real subject, 
\emph{`en fågel'} the one of an object.

Transitive verbs may not be used like this
\enumsentence{\shortexnt{7}
{*Det & äter & en & fågel & frön & på & taket}
{it & eats & a & bird & seeds & on & the roof}}

unless their in passive form
\enumsentence{\shortex{5}
{Det&dricks&mycket & öl& nuförtiden}
{It&drink+s&much & beer& nowadays}
{A lot of beer is being drunk these days}}
A very common example of this is sentences with the verbs \emph{finnas},
(\emph{exist})
\emph{saknas} (\emph{miss}) and \emph{fattas}(\emph{lack}).
\enumsentence{\shortex{3}
{Det & finns & kaffe}
{It & exist & coffee}
{''There is coffee"}}
\enumsentence{\shortex{3}
{Det & saknas&  kaffe}
{It & misses & coffee}
{''There is no coffee"}}
\enumsentence{\shortex{3}
{Det & fattas & kaffe}
{It & lacks & coffee}
{''There is no coffee"}}

A special GF category, \verb|SimpleVP|, was needed for verb phrases of the type
described above.
There are also restrictions on the real subject.
\enumsentence{\begin{tabular}{ll}
*Det sitter den fågeln på taket. & *Det sitter fågeln på taket.\\
``There are that bird sitting on the roof'' & ``There are the bird sitting on the roof'' \\
\end{tabular}}
Expressed in GF terms, the determiner must have definitness DIndef. That is, all determiners
that requireds the noun and the adjective to be used in indefinite form may be used.
%The quantifier must have determiner must havenoun must be in indefinite form,
%even though some determiners are allowed.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Det sitter några fåglar på taket    & & b. & *Det sitter fåglarna på taket\\
   & ``There are some birds on the roof''& &    & ~``There are the birds on the roof''\\
\end{tabular}}\label{sent:faglar}
Sentence \ref{sent:gamlafagel}b exemplifies a determiner that uses the noun in indefinite form,
but the adjective in definite form.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. & Det sitter en gammal fågel på taket.& & b. & * Det sitter min gamla fågel på taket.\\
   & ``There are some birds on the roof''& &    &~ ``There is my old bird on the roof.''\\
\end{tabular}}\label{sent:gamlafagel}

%%Quantifiers like \emph{`denna'} or determiners like \emph{`samtliga'} may
%%not be used, but the determiners \emph{`många'}, \emph{`några'} and the quantifier
%%\emph{`ingen'} work well.
%%Semantic difference, atm we allow any NP. \\
To form a clause with formal subject, we hence need to inspect the determiner.
\verb|FormalSub : SimpleVP -> Det -> CN -> Cl ;|\\
The function combines a verb phrase a determiner and a noun phrase in to a clause.
If the determiner requires a noun in determined form, the clause is put to NONEXIST.
However, this is not an ideal solution, but it cannot be seen on the type level
which definiteness a noun phrase has and we therefore need to do it at the runtime? level.

% move to future?
As a future direction it would be intresting to examine if the grammars noun phrases
should have more information on the type level. In the reflexive objects are added in (see p.
\pageref{swe:refl}) dependent types are used for showing if a phrase can be used as
subject. We would also like to produce the sentences, where `av' should be used
only when the noun has an explicit article or determiner.
\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}}
a. &De flesta hästarna  && b.& De flesta \textbf{av} de där hästarna.\\
 &``Most of the horses"  && b.&  ``Most of those horses."\\
\end{tabular}}\label{sent:mosthorses}

%There are other examples of when we need to know the definiteness of the noun phrase 
%%, for example when using %the reflexive possessive pronoun `sin' (see section `Reflexive pronouns' below)
%or when using some predeterminers (see section `Quantifiers'). Not true, not the same :(

@@

\textbf{Genitive}
the extramodule contained GenNP for costructing genitvives. In Swedish
done by adding s to the end of the word. In Gf, Uses the NPPoss form of the noun. 
severely overgenerating when started, and had to be restricted. Cycle:
GenNP : NP -> Quant, DetQuant : Quant -> Det; DetNP : Det -> NP
      katt -> kattens      ->          kattens           det är kattens
Restricted by removing the sp-field which is used by DetNP. Disallows ex 'det or kattens',
but still much better then the overgeneration.
Quants have no gen-form. min -> mins?? but when NPs they must have! otherwise min could be
min with GenNP applied n number of times. solutions: bind s.
or: A field should also been added to Quants, so that they can have a gen-form
 : somligas.
1 cannot be alone (fixed) 2 some words have id. (no) gen form (fixed)
           fixing this include restructuring all???

\textbf{Det-utr}. no effect for the some Dets, like 'sådan', no difference between utrum and neutrum.
 Can vary in number, since is a NP. Do not want to cut connection to abstract but changing them,
 so added one more. Leads to ambiguities in some cases, but considering that we would want to use
 this for translation, neutr and utr may differ in the other language.

@@

here or in robust: SupCl, det as VP or VV or S for S (remember Foc)
\subsubsection{Reflexive pronouns}
%varandra and sin/sitt/sig.
An important piece in a Swedish grammar is the reflexive pronouns and
%both \emph{sin} and , eg. \emph{sin}.
the possessive pronouns as described in p. \pageref{swe:refl}.
Those make up noun phrases which are not allowed in subjects, 
and we want to be able to parse to following sentences:

\enumsentence{Han såg sina barns skor.\\ \label{ex:ReflGen}
He saw \sc{self's} children's shoes.}
\enumsentence{Sina vantar hade han glömt på tåget.\\
\sc{self's} gloves, he had forgotten on the train.}
\enumsentence{Han ger sina pengar till sina syskon.\\
He gives \sc{self's} money to \sc{self's} children.}
\enumsentence{Han är längre än sin kompis.\\
He is taller than \sc{self's} friend.} \label{ex:ReflAP}
\vspace{-2mm}
\enumsentence{Han är här oftare än sin kompis.\\
He is here more often than \sc{self's} friend.} \label{ex:ReflAdv}
\vspace{-2mm}
\enumsentence{Hon tyckte om skolan och alla sina elever.\\
She liked the school and all \sc{self's} students.}\label{ex:ReflConj}
\enumsentence{Han såg sina få böcker och sin penna.\\
He saw \sc{self's}his few books and \sc{self's} pencil.}\label{ex:ReflConj2}
\vspace{-2mm}
\enumsentence{Hon ber alla sina kompisar att gå\\
She asks all \sc{self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave \sc{self's}.}
but not
\enumsentence{*Sina vantar var kvar på tåget.\\
\sc{self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Hon och sin kompis läser en bok.\\
She and \sc{self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give \sc{self's} money to \sc{self's}his siblings. }
\enumsentence{*Han vill att jag rakar sig.\\
He wants me to shave \sc{self's}.}

@@

The grammar has been extended to accept
\enumsentence{\shortex{7}
{\textbf{De uppfattningar}& som& förs& fram ...} 
{The opinions& that& put & forward ...}
{``The opinions presented ..."}}
where the definite artcile \emph{``de"} is used together with a indefinite
noun. This is only correct for the definite articles and not determiners
and requires that the noun phrase is followud by a restrictive
relative clauses or together with the verb \emph{``finnas"}.
In other cases definitev form must be used, cf. sentence \ref{sent:debocker}.
\enumsentence{\textbf{De uppfattningarna} förs fram.\\
``The opinions are presented."}\label{sent:debocker}
% also ok in finnas!!

@@

% compared to penn treebank: 82 tags. talbanken: at least 130
% 99.1 % of Penn sentences have 6 or fewer clauses

@@

\section{Future Work}
\label{sec:future}
>> divide into consequences and future plans.
What is still left to do? Description and direction of future work:
Handling unknown words and , handling unknown grammar, 
how to make the parser robust, ellipses and long sentences, names and numbers.
What is still needed in the grammar, how important that is. 
\subsection{Grammar}
köpa bil 
object shift
\subsection{Parallel parsing/Chunk parsing}
\subsection{Lexicon}
compounds
Speed up! : perhaps have one lexicon for morph. analyse and one for parsing.
exchange words.
%The list of not-imported can be used for adding other kinds of lexicons - idioms, have pos-tags.
%Also to analyse what's wrong, or use the words with a word guesser. The code
%has been adapted for this.
%For exapmle pronouns - dets, have combined information from Talbanken to add them in the correct
%category. How this works, look at tags. Small amount, can be done manually.
